---
title: "R Notebook"
output: html_notebook
---

# Libraries 

```{r}
library(tidyverse)
library(here)
library(maps)
library(mapdata)
library(kableExtra)

#library(mapproj)
#library(raster)
#library(viridis)
#library(leaflet)
```


# Load Data 

```{r}
US_causes_of_death <- read_csv(here("Final_Project", "data", "Leading_Causes_of_Death_United_States.csv"))

US_death <- US_causes_of_death %>%
  select(Year, `Cause Name`, State, Deaths)

view(US_death)



states<-map_data("state")
view(states)
```



# Manipulate 

```{r}
US <- US_death %>%
  mutate(
    region = case_when( #recode the State Names to be Capitalized
      State == "Alabama" ~ "alabama",
      State == "Arizona" ~ "arizona",
      State == "Arkansas" ~ "arkansas",
      State == "California" ~ "california",
      State == "Colorado" ~ "colorado",
      State == "Connecticut" ~ "connecticut",
      State == "Delaware" ~ "delaware",
      State == "Florida" ~ "florida",
      State == "Georgia" ~ "georgia",
      State == "Idaho" ~ "idaho",
      State == "Illinois" ~ "illinois",
      State == "Indiana" ~ "indiana",
      State == "Iowa" ~ "iowa",
      State == "Kansas" ~ "kansas",
      State == "Kentucky" ~ "kentucky",
      State == "Louisiana" ~ "louisiana",
      State == "Maine" ~ "maine",
      State == "Maryland" ~ "maryland",
      State == "Massachusetts" ~ "massachusetts",
      State == "Michigan" ~ "michigan",
      State == "Minnesota" ~ "minnesota",
      State == "Mississippi" ~ "mississippi",
      State == "Missouri" ~ "missouri",
      State == "Montana" ~ "montana",
      State == "Nebraska" ~ "nebraska",
      State == "Nevada" ~ "nevada",
      State == "New Hampshire" ~ "new hampshire",
      State == "New Jersey" ~ "new jersey",
      State == "New Mexico" ~ "new mexico",
      State == "New York" ~ "new york",
      State == "North Carolina" ~ "north carolina",
      State == "North Dakota" ~ "north dakota",
      State == "Ohio" ~ "ohio",
      State == "Oklahoma" ~ "oklahoma",
      State == "Oregon" ~ "oregon",
      State == "Pennsylvania" ~ "pennsylvania",
      State == "Rhode Island" ~ "rhode island",
      State == "South Carolina" ~ "south carolina",
      State == "South Dakota" ~ "south dakota",
      State == "Tennessee" ~ "tennessee",
      State == "Texas" ~ "texas",
      State == "Utah" ~ "utah",
      State == "Vermont" ~ "vermont",
      State == "Virginia" ~ "virginia",
      State == "Washington" ~ "washington",
      State == "West Virginia" ~ "west virginia",
      State == "Wisconsin" ~ "wisconsin",
      State == "Wyoming" ~ "wyoming",
      State == "District of Columbia" ~ "washington"))


# Filter for the years 1999 and 2017

stat_death_1999 <- US %>%
   filter(complete.cases(.),
          Year == 1999,
#maybe remove #`Cause Name` != "All causes"
          ) %>%
  select(Year, 'Cause Name', Deaths, region)

stat_death_2017 <- US %>%
   filter(complete.cases(.),
          Year == 2017) %>%
  select(Year, 'Cause Name', Deaths, region)


view(stat_death_1999)
view(stat_death_2017)
```



# Join the data 

```{r}
# Data from 1999
mapdata_1999 <- (full_join(states, stat_death_1999))


# Data from 2017
mapdata_2017 <- (full_join(states, stat_death_2017))

view(mapdata_1999)

view(mapdata_2017)
```


# Map the data

```{r}
map1 <- mapdata_1999 %>%
  ggplot(aes(x= long, 
             y = lat, 
             map_id = region)) +
  geom_polygon(aes(fill = Deaths), 
               color = "Black")+
  coord_map("ortho", orientation = c(39, -98, 0))+
  scale_fill_gradient(trans = "log10")+
  labs(title = "Leading Causes of Death is the United States",
       subtitle = "Fiscal Years 1999 - 2017",
       caption = "Source: Centers for Disease Control and Prevention. Data.gov")+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        rect = element_blank(),
        plot.title = element_text(size  =24, face = "bold"))
  

map1


```


Testing New things 
```{r}
map2 <- mapdata_1999 %>%
  group_by("Cause Name") %>%
  ggplot(aes(x= long, 
             y = lat, 
             map_id = region)) +
  geom_polygon(aes(x = long,
                   y = lat,
                   fill = Deaths, 
               color = "Cause Name"))+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        rect = element_blank())+
  coord_map("ortho", orientation = c(39, -98, 0))+
  labs(title = "Leading of Causes of Death is the United States")
  

map2
```


```{r}
map3 <- mapdata_1999 %>%
  group_by("Cause Name") %>%
  ggplot(aes(x= long, 
             y = lat, 
             map_id = region)) +
  geom_polygon(aes(x = long,
                   y = lat,
                   fill = "Cause Name"), 
               color = "Black")+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        rect = element_blank())+
  coord_map("ortho", orientation = c(39, -98, 0))+
  labs(title = "Leading of Causes of Death is the United States")
  

map3
```






```{r}
HD_1999 <- mapdata_1999 %>%
  filter(`Cause Name` == "Heart disease")

view(HD_1999)

map4 <- HD_1999 %>%
  ggplot(aes(x= long, 
             y = lat, 
             map_id = region)) +
  geom_polygon(aes(fill = Deaths), 
               color = "Black")+
  coord_map("ortho", orientation = c(39, -98, 0))+
  scale_fill_gradient(trans = "log10")+
  labs(title = "Heart disease - related deaths in the United States",
       subtitle = "Fiscal Year 1999",
       caption = "Source: Centers for Disease Control and Prevention. Data.gov")+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        rect = element_blank(),
        plot.title = element_text(size  =24, face = "bold"))
  

map4
```



```{r}
map5 <- mapdata_1999 %>%
  ggplot(aes(x= long, 
             y = lat, 
             group = `Cause Name`)) +
  geom_map(
    aes(map_id = region),
    map = states,
    color = "gray80", 
    fill = "gray30",
    size = 0.3
  )+
  coord_map("ortho", orientation = c(39, -98, 0))+
  geom_polygon(aes(group = group,
                   fill = Deaths), 
               color = "Black")+
  scale_fill_gradient(trans = "log10")+
  labs(title = "Leading of Causes of Death is the United States")+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        rect = element_blank(),
        plot.title = element_text(size  =24, face = "bold"),
        legend.position = "bottom")
  

map5
```














































### Might need later
region == "alaska" ~ "Alaska",
region == "hawaii" ~ "Hawaii"))


states <- map_data("state") %>%
  mutate(
    State = case_when( #recode the State Names to be Capitalized
      region == "alabama" ~ "Alabama",
      region == "arizona" ~ "Arizona",
      region == "arkansas" ~ "Arkansas",
      region == "california" ~ "California",
      region == "colorado" ~ "Colorado",
      region == "connecticut" ~ "Connecticut",
      region == "delaware" ~ "Delaware",
      region == "florida" ~ "Florida",
      region == "georgia" ~ "Georgia",
      region == "idaho" ~ "Idaho",
      region == "illinois" ~ "Illinois",
      region == "indiana" ~ "Indiana",
      region == "iowa" ~ "Iowa",
      region == "kansas" ~ "Kansas",
      region == "kentucky" ~ "Kentucky",
      region == "louisiana" ~ "Louisiana",
      region == "maine" ~ "Maine",
      region == "maryland" ~ "Maryland",
      region == "massachusetts" ~ "Massachusetts",
      region == "michigan" ~ "Michigan",
      region == "minnesota" ~ "Minnesota",
      region == "mississippi" ~ "Mississippi",
      region == "missouri" ~ "Missouri",
      region == "montana" ~ "Montana",
      region == "nebraska" ~ "Nebraska",
      region == "nevada" ~ "Nevada",
      region == "new hampshire" ~ "New Hampshire",
      region == "new jersey" ~ "New Jersey",
      region == "new mexico" ~ "New Mexico",
      region == "new york" ~ "New York",
      region == "north carolina" ~ "North Carolina",
      region == "north dakota" ~ "North Dakota",
      region == "ohio" ~ "Ohio",
      region == "oklahoma" ~ "Oklahoma",
      region == "oregon" ~ "Oregon",
      region == "pennsylvania" ~ "Pennsylvania",
      region == "rhode island" ~ "Rhode Island",
      region == "south carolina" ~ "South Carolina",
      region == "south dakota" ~ "South Dakota",
      region == "tennessee" ~ "Tennessee",
      region == "texas" ~ "Texas",
      region == "utah" ~ "Utah",
      region == "vermont" ~ "Vermont",
      region == "virginia" ~ "Virginia",
      region == "washington" ~ "Washington",
      region == "west virginia" ~ "West Virginia",
      region == "wisconsin" ~ "Wisconsin",
      region == "wyoming" ~ "Wyoming",
      region == "washington" ~ "District of Columbia"))
  

view(states)

st_map <- states %>%
  select(long, lat, group, order, State)

view(st_map)















































































# Create Figues for leading causes of death


## Libraries 
```{r}
library(tidyverse)
library(here)
library(dplyr)
```


## Manipulate Data

```{r}
US_causes_of_death <- read_csv(here("Final_Project", "data", "Leading_Causes_of_Death_United_States.csv"))

US_DEATH <- US_causes_of_death %>%
  select(Year, `Cause Name`, State, Deaths) %>%
  filter(State == "United States") 

view(US_DEATH)

```



# Plot 

```{r}
Top10 <- US_DEATH %>% 
  filter(`Cause Name`!= "All causes") %>%
  ggplot( aes(x= Year, y= Deaths, group= `Cause Name`, color= `Cause Name`)) +
    geom_line()+
  labs(title = "Leading Causes of Death in the United States")+
  theme_classic()+
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank())

Top10

```

```{r}
Eight <- US_DEATH %>% 
  filter(`Cause Name` != "Cancer",
         `Cause Name` != "Heart disease",
         `Cause Name`!= "All causes")

Top8 <- Eight %>%
  ggplot( aes(x= Year, y= Deaths, group= `Cause Name`, color= `Cause Name`)) +
    geom_line()+
  theme_classic()+
     theme(legend.position = "none")


Top8
```



# Patchwork to bring the line plots together

```{r}
library(patchwork)

######## NEED TO MAKE SURE THAT THE COLORS ARE MATCHING IN THE PATCHWORK #########

Top10/ Top8 +
  plot_layout(guides = 'collect')
```



























### Might need later
region == "alaska" ~ "Alaska",
region == "hawaii" ~ "Hawaii"))


states <- map_data("state") %>%
  mutate(
    State = case_when( #recode the State Names to be Capitalized
      region == "alabama" ~ "Alabama",
      region == "arizona" ~ "Arizona",
      region == "arkansas" ~ "Arkansas",
      region == "california" ~ "California",
      region == "colorado" ~ "Colorado",
      region == "connecticut" ~ "Connecticut",
      region == "delaware" ~ "Delaware",
      region == "florida" ~ "Florida",
      region == "georgia" ~ "Georgia",
      region == "idaho" ~ "Idaho",
      region == "illinois" ~ "Illinois",
      region == "indiana" ~ "Indiana",
      region == "iowa" ~ "Iowa",
      region == "kansas" ~ "Kansas",
      region == "kentucky" ~ "Kentucky",
      region == "louisiana" ~ "Louisiana",
      region == "maine" ~ "Maine",
      region == "maryland" ~ "Maryland",
      region == "massachusetts" ~ "Massachusetts",
      region == "michigan" ~ "Michigan",
      region == "minnesota" ~ "Minnesota",
      region == "mississippi" ~ "Mississippi",
      region == "missouri" ~ "Missouri",
      region == "montana" ~ "Montana",
      region == "nebraska" ~ "Nebraska",
      region == "nevada" ~ "Nevada",
      region == "new hampshire" ~ "New Hampshire",
      region == "new jersey" ~ "New Jersey",
      region == "new mexico" ~ "New Mexico",
      region == "new york" ~ "New York",
      region == "north carolina" ~ "North Carolina",
      region == "north dakota" ~ "North Dakota",
      region == "ohio" ~ "Ohio",
      region == "oklahoma" ~ "Oklahoma",
      region == "oregon" ~ "Oregon",
      region == "pennsylvania" ~ "Pennsylvania",
      region == "rhode island" ~ "Rhode Island",
      region == "south carolina" ~ "South Carolina",
      region == "south dakota" ~ "South Dakota",
      region == "tennessee" ~ "Tennessee",
      region == "texas" ~ "Texas",
      region == "utah" ~ "Utah",
      region == "vermont" ~ "Vermont",
      region == "virginia" ~ "Virginia",
      region == "washington" ~ "Washington",
      region == "west virginia" ~ "West Virginia",
      region == "wisconsin" ~ "Wisconsin",
      region == "wyoming" ~ "Wyoming",
      region == "washington" ~ "District of Columbia"))
  

view(states)

st_map <- states %>%
  select(long, lat, group, order, State)



























# Create a Pie Plot of the Top Three preventable Causes of death

```{r}
# Figure out what what to divide by to calculate the percentages 
all_causes <- US_causes_of_death %>%
  select(Year, `Cause Name`, State, Deaths) %>%
  filter(State == "United States",
         `Cause Name` == "All causes",
         Year %in% c(1999, 2017))

view(all_causes)


# Create table for the top preventable causes of death
PI_death <- US_causes_of_death %>%
  select(Year, `Cause Name`, State, Deaths) %>%
  filter(State == "United States",
         `Cause Name` %in% c("CLRD", "Heart disease", "Diabetes", "All causes"))  # Start by filtering out the Heart Disease, Diabetes, Chronic lower respiratory disease
  

view(PI_death)
```


```{r}
# Filter for earliest time period 
Pie_1999 <- PI_death %>% 
  filter(`Year` == 1999)


per_1999 <- Pie_1999 %>%
  mutate(
    Percent_Death = case_when(
      `Cause Name` == "CLRD" ~ 100*124181/2391399,
      `Cause Name` == "Diabetes" ~ 100*68399/2391399,
      `Cause Name` == "Heart disease" ~ 100*725192/2391399
      `Cause Name` == "all causes" ~ 100*725192/2391399
    )
  )

view(per_1999)



# Filter for most recent time period 
Pie_2017 <- PI_death %>% 
  filter(`Year` == 2017)

view(Pie_2017)


per_2017 <- Pie_2017 %>%
  mutate(
    Percent_Death = case_when(
      `Cause Name` == "CLRD" ~ 100*160201/2813503,
      `Cause Name` == "Diabetes" ~ 100*83564/2813503,
      `Cause Name` == "Heart disease" ~ 100*647457/2813503
      `Cause Name` == ""
    )
  )

view(per_2017)
```



## Plot Pie charts

```{r}
per_1999 %>%
  pie(Percent_Death)
```

