---
title: "Causes of Death in the US"
author: "Anthony Gutierrez"
date: "5/4/2021"
output:
  html_document:
    df_print: paged
  word_document: default
---

# Libraries 

```{r}
library(tidyverse)
library(here)
library(patchwork)
library(RColorBrewer)
library(kableExtra)

library(maps)
library(mapdata)
library(mapproj)

library(cowsay)
library(beepr)
library(praise)

```

WELCOME!

```{r}
say("Welcome", by = "hypnotoad")
```


# Introduction

What are the leading causes of death in the United States? 

Findings:

* Preventable 
* Unhealthy lifestyle choices 
* Mapping showed increase in preventable causes of death



# Load in Data

```{r}
# This is the primary source of data:
US_causes_of_death <- read_csv(here("data", "Leading_Causes_of_Death_United_States.csv"))


# This is a weekly count of deaths. I will use this to plot COVID deaths
US_weekly_death <- read_csv(here("data", "Weekly_Counts_of_Deaths.csv"))




# The following datasets may explain why we are seeing a significant amount of certain deaths:

Obesity_health <- read_csv(here("data","Nutrition_Physical_Activity_and_Obesity.csv"))

Tobacco_consumption <- read_csv(here("data", "Adult_Tobacco_Consumption.csv"))

```


# What are the leading causes of Death in the United States?

## Manipulate the data

```{r}
US_DEATH <- US_causes_of_death %>%
  select(Year, `Cause Name`, State, Deaths) %>% #select for columns involving the listed
  filter(State == "United States") # filter for only national data (filter out the state - level data)

```

## Plot 

```{r}
# Create line plot
Top10 <- US_DEATH %>% 
  filter(`Cause Name`!= "All causes") %>% # remove all causes so that it does not skew the data
  ggplot( aes(x= Year, y= Deaths, group= `Cause Name`, color= `Cause Name`)) + # plot number of deaths by year for each cause
    geom_line()+ #create line plot for each of the top 10 reasons for death in the US
  labs(title = "Leading Causes of Death in the United States")+
  theme_classic()+
  theme(axis.text.x = element_blank(), #remove x- axis text and title because I will patch with another figure later
        axis.title.x = element_blank())

Top10
```

```{r}

Eight <- US_DEATH %>% 
  filter(`Cause Name` != "Cancer",
         `Cause Name` != "Heart disease",
         `Cause Name`!= "All causes") #Filter out all causes and the top two causes of death to see the remaining line plots with greater clarity

#create second line plot
#exclude figure title and legend title, include x-axis
Top8 <- Eight %>%
  ggplot( aes(x= Year, y= Deaths, group= `Cause Name`, color= `Cause Name`)) +
    geom_line()+   
  theme_classic()+
     theme(legend.title = element_blank())

```



# Patchwork to bring the line plots together

```{r}
Top10/ Top8 +
  plot_layout(guides = 'collect')+
   plot_annotation(subtitle = "Source: Centers for Disease Control and Prevention")

# Need two legends because I could not get the color coding to match up
```


# Create a Pie Plot of the Top Three preventable Causes of death

## Manipulate Data

```{r}
PI_death <- US_DEATH %>%
  filter(`Cause Name` == c("All causes","Heart disease", "CLRD", "Diabetes")) #filter out the leading preventable causes of death

```


```{r}
# Filter for earliest time period 
Pie_1999 <- US_DEATH %>% 
  filter(`Year` == 1999) 

per_1999 <- Pie_1999 %>%
  mutate( #create a column for percentage of dead from that cause
    Percent_Death = case_when(
      `Cause Name` == "CLRD" ~ 100*124181/2391399, #people that died of the disease / total number of people that died that year
      `Cause Name` == "Diabetes" ~ 100*68399/2391399,
      `Cause Name` == "Heart disease" ~ 100*725192/2391399,
      `Cause Name` == "All causes" ~ 61.621962))


# Filter for most recent time period 
Pie_2017 <- US_DEATH %>% 
  filter(`Year` == 2017)

per_2017 <- Pie_2017 %>%
  mutate(
    Percent_Death = case_when(
      `Cause Name` == "CLRD" ~ 100*160201/2813503,
      `Cause Name` == "Diabetes" ~ 100*83564/2813503,
      `Cause Name` == "Heart disease" ~ 100*647457/2813503,
      `Cause Name` == "All causes" ~ 68.323403))

```


## Create color palette
```{r}
myPalette <- brewer.pal(4, "Set2") 
```


## Assign values to the pi slices 

```{r}
# 1999 Data 
per_1999 %>%
  kbl() %>% # make it a kable table to see all my percentages easily
  kable_classic() 

a <- c(61.62, 2.86, 5.19, 30.33)
```

```{r}
# 2017 Data
per_2017 %>%
  kbl() %>% # make it a kable table
  kable_classic()

b <- c(68.32, 2.97, 5.69, 23.01)
```


## Plot pi charts

```{r}
pie(a, 
    clockwise = TRUE,
    labels = c("Other", "Diabetes", "CLRD", "Heart disease"),
    border="white", 
    col=myPalette ) 
```

```{r}
pie(b, 
    clockwise = TRUE,
    labels = c("Other", "Diabetes", "CLRD", "Heart disease"),
    border="white", 
    col=myPalette ) 

# I was unable to patch pi charts together because they are not ggplots. ggplotify did not work
```


Discussion:

Decrease in Heart disease may be related in the more common use of and emergence of new antihypertensive drugs (drugs that reduce blood pressure). Plot from earlier also showed a decreased in stroke - related death.

I was curious to know why respiratory disease death increased even though tobacco consumption has been dropping




# Possible explanations

Explanations for why our heart disease deaths are generally high and why CLRD- associated deaths did not decrease over the past two decades : Tobacco consumption and Obesity 

## Manipulate data

```{r}
#Tobacco consumption dataset
Yearly_con <- Tobacco_consumption %>%
  filter(Year == c(2000, 2019), #Look at earliest and most recent years
         Measure == "All Combustibles") # look at all combustible (smokeble) tobacco products that were sold



# Obesity dataset
Health <- Obesity_health %>%
  filter(LocationDesc == "National",
         Question == c("Percent of adults aged 18 years and older who have obesity", "Percent of adults aged 18 years and older who have an overweight classification"), #filter for obesity and overweight 
         StratificationCategory1 == "Age (years)", #only age, ignore race, education, etc. 
         `Age(years)` == "18 - 24") %>%
  select(1:11)

```


## Plot Tobacco consumption

```{r}
Yearly_con %>%
  group_by(Measure) %>%
  ggplot(aes(x = Year,
             y = Total))+
  geom_col() +
  labs(title = "Total Tobacco consumption in the United States",
       subtitle = "Fiscal years 2000 and 2019",
       y = "Cigarette Equivalents")

```
Possible explanation: People from older generations are dying. 


## Plot percent of overweight and obese persons in the US
```{r}
Health %>%
  ggplot(aes(x = YearStart,
             y = Data_Value,
             group = Question, #obese or overweight
             color = Question))+
  geom_line()+
  labs(title = "Percent of US Population that is Overweight or Obese",
       caption = "Source: Data.gov. Centers for Disease Control and Prevention",
       x = "Year",
       y = "Percentage")+
  theme_classic()+
  theme(legend.position = "bottom",
        legend.direction = "vertical")
  
```



# Map heart disease, CLRD, & diabetes deaths

## Manipulate the data
```{r}
US_death <- US_causes_of_death %>%
  select(Year, `Cause Name`, State, Deaths)

# Load in mapping data
states<-map_data("state")

```


```{r}
# To join the US_death data to the state mapping data, I need to first recode the state names under the state column of the death dataset to match the state names under region column of the state dataset.

US <- US_death %>%
  mutate(
    region = case_when( #recode the State Names to be Capitalized
      State == "Alabama" ~ "alabama",
      State == "Arizona" ~ "arizona",
      State == "Arkansas" ~ "arkansas",
      State == "California" ~ "california",
      State == "Colorado" ~ "colorado",
      State == "Connecticut" ~ "connecticut",
      State == "Delaware" ~ "delaware",
      State == "Florida" ~ "florida",
      State == "Georgia" ~ "georgia",
      State == "Idaho" ~ "idaho",
      State == "Illinois" ~ "illinois",
      State == "Indiana" ~ "indiana",
      State == "Iowa" ~ "iowa",
      State == "Kansas" ~ "kansas",
      State == "Kentucky" ~ "kentucky",
      State == "Louisiana" ~ "louisiana",
      State == "Maine" ~ "maine",
      State == "Maryland" ~ "maryland",
      State == "Massachusetts" ~ "massachusetts",
      State == "Michigan" ~ "michigan",
      State == "Minnesota" ~ "minnesota",
      State == "Mississippi" ~ "mississippi",
      State == "Missouri" ~ "missouri",
      State == "Montana" ~ "montana",
      State == "Nebraska" ~ "nebraska",
      State == "Nevada" ~ "nevada",
      State == "New Hampshire" ~ "new hampshire",
      State == "New Jersey" ~ "new jersey",
      State == "New Mexico" ~ "new mexico",
      State == "New York" ~ "new york",
      State == "North Carolina" ~ "north carolina",
      State == "North Dakota" ~ "north dakota",
      State == "Ohio" ~ "ohio",
      State == "Oklahoma" ~ "oklahoma",
      State == "Oregon" ~ "oregon",
      State == "Pennsylvania" ~ "pennsylvania",
      State == "Rhode Island" ~ "rhode island",
      State == "South Carolina" ~ "south carolina",
      State == "South Dakota" ~ "south dakota",
      State == "Tennessee" ~ "tennessee",
      State == "Texas" ~ "texas",
      State == "Utah" ~ "utah",
      State == "Vermont" ~ "vermont",
      State == "Virginia" ~ "virginia",
      State == "Washington" ~ "washington",
      State == "West Virginia" ~ "west virginia",
      State == "Wisconsin" ~ "wisconsin",
      State == "Wyoming" ~ "wyoming",
      State == "District of Columbia" ~ "washington"))


# Filter for the years 1999 and 2017, respectively 

stat_death_1999 <- US %>%
   filter(complete.cases(.), # removes all rows that have NAs
          Year == 1999) %>% #only look at 1999 data
  select(Year, 'Cause Name', Deaths, region) #select for these columns

stat_death_2017 <- US %>%
   filter(complete.cases(.), # removes all rows that have NAs
          Year == 2017) %>% #only look at 2017 data
  select(Year, 'Cause Name', Deaths, region) #select for these columns

```



## Join the state(mapping) data and the death statistics 

```{r}
# Data from 1999
mapdata_1999 <- (full_join(states, stat_death_1999)) #Join the cleaned up US_death data to the state mapping data [1999]


# Data from 2017
mapdata_2017 <- (full_join(states, stat_death_2017)) #Join the cleaned up US_death data to the state mapping data [2017]


#NOTE: all of this is prep to map the three leading causes of death individually
```


## Map the data

### Map heart disease deaths for 1999 and 2017
```{r}
# Fix data for 1999
HD_1999 <- mapdata_1999 %>%
  filter(`Cause Name` == "Heart disease") #filter for heart disease deaths


HD_map99 <- HD_1999 %>%
  ggplot(aes(x= long, #plot lat and long 
             y = lat, 
             map_id = region)) +
  geom_polygon(aes(fill = Deaths), #create polygons for the shapefiles and fill with number of deaths
               color = "Black")+
  coord_map("ortho", orientation = c(39, -98, 0))+ #"NEVER USE UNPROJECTED DATA!!" -Sensai Rachman
  scale_fill_gradient(trans = "log10", low = "white", high = "dark blue")+
  labs(subtitle = "Fiscal Year 1999")+
  theme(axis.text.x = element_blank(), #got rid of axis's because they are unnecessary 
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        rect = element_blank(),
        legend.position = "none")

```


```{r}
# Fix data for 2017
HD_2017 <- mapdata_2017 %>%
  filter(`Cause Name` == "Heart disease") #filter for heart disease deaths


HD_map17 <- HD_2017 %>%
  ggplot(aes(x= long, 
             y = lat, 
             map_id = region)) +
  geom_polygon(aes(fill = Deaths),  
               color = "Black")+
  coord_map("ortho", orientation = c(39, -98, 0))+ #
  scale_fill_gradient(trans = "log10", low = "white", high = "dark blue")+
  labs(subtitle = "Fiscal year 2017")+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        rect = element_blank())

```

###  Bring HD plots together with Patchwork
```{r}
HD_map99 + HD_map17 + #patch the HD maps together and assign a title and caption
  plot_annotation(title = 'Heart Disease - Related Deaths \nUnited States',
                  caption = "Source: Centers for Disease Control and Prevention. Data.gov")+
  theme(plot.title = element_text(size = 24, face = "bold"))
```




### Map Cronic lower respiratory disease deaths for 1999 and 2017

```{r}
#Fix data for 1999
CLRD_1999 <- mapdata_1999 %>%
  filter(`Cause Name` == "CLRD") #filter for CLRD


CLRD_map99 <- CLRD_1999 %>%
  ggplot(aes(x= long, 
             y = lat, 
             map_id = region)) + # plot states
  geom_polygon(aes(fill = Deaths), #color depth depends on number of deaths
               color = "Black")+
  coord_map("ortho", orientation = c(39, -98, 0))+ #project in ortho and orient to show only the US in best position
  scale_fill_gradient(trans = "log10", low = "yellow", high  = "red")+ #color gradient
  labs(subtitle = "Fiscal Year 1999")+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        rect = element_blank(), #remove axis's and background as they are unnecessary 
        legend.position = "none") 

```


```{r}
#Fix data for 2017
CLRD_2017 <- mapdata_2017 %>%
  filter(`Cause Name` == "CLRD") #filter for CLRD


CLRD_map17 <- CLRD_2017 %>%
  ggplot(aes(x= long, 
             y = lat, 
             map_id = region)) + # plot states
  geom_polygon(aes(fill = Deaths), #color depth depends on number of deaths
               color = "Black")+
  coord_map("ortho", orientation = c(39, -98, 0))+ #project in ortho
  scale_fill_gradient(trans = "log10", low = "yellow", high = "red")+ #color gradient
  labs(subtitle = "Fiscal Year 2017")+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        rect = element_blank()) #remove axis's and background as they are unnecessary 

```


### Bring CLRD plots together with Patchwork

```{r}
CLRD_map99 + CLRD_map17 +
  plot_annotation(title = 'Chronic Lower Respritory Disease - Related Deaths \nUnited States',
                  caption = "Source: Centers for Disease Control and Prevention. Data.gov")+
  theme(plot.title = element_text(size = 24, face = "bold"))
```





### Map diabetes deaths for 1999 and 2017

```{r}
#Fix data for 1999
Dia_1999 <- mapdata_1999 %>%
  filter(`Cause Name` == "Diabetes") #filter for diabetes


Dia_map99 <- Dia_1999 %>%
  ggplot(aes(x= long, 
             y = lat, 
             map_id = region)) + # plot states
  geom_polygon(aes(fill = Deaths), #color depth depends on number of deaths
               color = "Black")+
  coord_map("ortho", orientation = c(39, -98, 0))+ #project in ortho
  scale_fill_gradient(trans = "log10", low = "pink", high  = "purple")+ #color gradient
  labs(subtitle = "Fiscal Year 1999")+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        rect = element_blank(), #remove axis's and background as they are unnecessary 
        legend.position = "none")

```

```{r}
#Fix data for 2017
Dia_2017 <- mapdata_2017 %>%
  filter(`Cause Name` == "Diabetes") #filter for diabetes


Dia_map17 <- Dia_2017 %>%
  ggplot(aes(x= long, 
             y = lat, 
             map_id = region)) + # plot states
  geom_polygon(aes(fill = Deaths), #color depth depends on number of deaths
               color = "Black")+
  coord_map("ortho", orientation = c(39, -98, 0))+ #project in ortho
  scale_fill_gradient(trans = "log10", low = "pink", high  = "purple")+ #color gradient
  labs(subtitle = "Fiscal Year 2017")+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        rect = element_blank()) #remove axis's and background as they are unnecessary 

```




###  Bring HD plots together with Patchwork

```{r}
Dia_map99 + Dia_map17 +
  plot_annotation(title = 'Diabetes - Related Deaths \nUnited States',
                  caption = "Source: Centers for Disease Control and Prevention. Data.gov")+
  theme(plot.title = element_text(size = 24, face = "bold"))
```



###  Bring all plots together with Patchwork
```{r}
HD_map17 + CLRD_map17 + Dia_map17 +
  plot_annotation(title = "Top Three Preventable Diseases in the United States",
                  subtitle = "Heart Disease, CLRD, & Diabetes",
                  caption = "Source: Centers for Disease Control and Prevention. Data.gov")+
  theme(plot.title = element_text(size = 24, face = "bold", color = "red"))
```



# "Easy" tips to live a healthier life
1.) Sleep
2.) No sugar in the morning
3.) Positive mentality





# FINISH

```{r}
# Time to celebrate!
beep(4) 
beep(8) 
praise()

```

